Intrusion Detection with \\ Genetic Algorithms and Fuzzy Logic

### 2013-11-26 Talk/Presentation.tex
Split the GA section into two slides, changed some things on the detection methodologies slide, also added some things to The Big Picture slide.

\begin{frame}
  \frametitle{The Big Picture}

  \begin{columns}
  \begin{column}{0.6\textwidth}
  \begin{itemize}
  	\item 
	\item have story here,
	\item also explain root somewhere
	\item 
	\item 
  \end{itemize}
  \end{column}
  \begin{column}{0.4\textwidth}

@@

\subsection{Detection Methodologies and Rules}
\begin{frame}
  \frametitle{Detection Methodologies}
  \begin{itemize}
  	\item Signature-based detection: compares well-known patterns of attacks
that are already known to the IDS against captured events in order to identify a possible attack.
	\begin{itemize}
		\item It is a simple and effective way to detect known attacks, but is ineffective against new kinds of unknown attacks.
	\end{itemize}
  	\item Anomaly-based detection: looks for patterns of activity that are rare and uncommon.
	\begin{itemize}
		\item It is harder to do than signature-based detection, but it can be an effective way to detect new, unknown attacks.
	\end{itemize}
  \end{itemize}
\end{frame}

@@

\subsection{Data Sets - KDD99 and RLD09}
\begin{frame}
  \frametitle{KDD99}
	\begin{itemize}
		\item Generated by simulating a military network environment in 1999.
		\item Has long been a standard data set for intrusion detection.
		\item Data was processed into 5 million records.
			\begin{itemize}
				\item A record is a sequence of TCP packets, between which data flows to and from a source IP address to a target IP address.
			\end{itemize}
		\item Data in the set is classified as normal or attack activity.
		\item Uses 41 features, which are properties of a record that are used to describe the activity and help to distinguish normal connections from attacks.

@@

\begin{frame}
  \frametitle{RLD09}
	\begin{itemize}
		\item RLD09 was created because KDD99 is 14 years old, and newer attack types are not included in it because of its age.
		\item Data was captured from a university in Bangkok, Thailand.
		\item 17 different types of attacks (divided into denial of service and probe attacks).

\linespace
\linespace
\linespace
\linespace
		\item .............maybe have something like "the authors of [6] used only the numerical features of KDD99 (34 out of 41)." if I decide to explain that paper.
	\end{itemize}
\end{frame}

@@

\subsection{Determining the Accuracy of an Algorithm}
\begin{frame}
  \frametitle{Determining the Accuracy of an Algorithm}
\centering{Predicted}
\begin{table}
Actual
\begin{tabular}{l|ll}
 & Not Attack & Attack \\ \hline
Not Attack & True Negative (TN) & False Positive (FP) \\
Attack & False Negative (FN) & True Positive (FP)  \\
\end{tabular}
\end{table}
\linespace
	\begin{itemize}
		\item Detection rate (DR): the percentage of normal and attack activity correctly classified from the total number of data records.

		\linespace
		\linespace
		\linespace
		\linespace
		\linespace
		\item .........maybe talk about the other paper's DR?
	\end{itemize}
\end{frame}

\subsection{Genetic Algorithms}
\begin{frame}
  \frametitle{Genetic Algorithms}
	\begin{itemize}
		\item GAs are a search technique used to find solutions to problems. Operations analogous to biological mutation, selection, and crossover are used to evolve and improve solutions.
        \item Possible solutions to problems can be represented in a variety of problem dependent ways, such as bit strings.
        \item First, a randomly generated population of potential solutions is created. Then mutation, crossover, and selection are applied to each generation until an acceptable solution is found or some time limit is exceeded.
        \item Mutation: random bits in an individual, or possible solution, are randomly changed.
        \item Crossover: Two individuals swap sequences of bits to form two new individuals.
        \item Selection: individuals that have better fitness are chosen to be parents.
        \item The fitness of an individual is specified by the fitness function, which determines the quality of a particular individual.
	\end{itemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Using Fuzzy Genetic Algorithms]{Using Fuzzy Genetic Algorithms}

@@

\begin{frame}
	\frametitle{Algorithm Overview}
	\begin{itemize}
		\item The algorithm generates rules and improves rules.
		\item One record is passed into a rule.
		\item Each feature in a record is matched to one block of the rule.
		\item The parameters of each block measure the degree of certainty of an attack using the trapezoidal fuzzy rule shape.
		\item The sum of the degrees of certainty from each block are then compared with a threshold to determine if the record represents an attack or normal behavior.
	\end{itemize}
\end{frame}

@@

\begin{frame}
	\frametitle{Experiments}
	\begin{itemize}
		\item A variety of experiments were run. Two experiments used just RLD09. Three experiments used both the RLD09 and KDD99 in order to compare how the fuzzy GA would perform on both.
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Experiments Using Only RLD09}
	Experiment 1
	\begin{itemize}
		\item Fuzzy GA was used to create DoS and probe detection rules. Both of the DoS and probe rules were then used together in the testing process to identify attacks from the testing data set.
		\item 10,000 records were used for the training set, 26,500 records were used for the test set.
	\end{itemize}

@@

	\bibitem{DBLP:journals/corr/abs-1204-1336}
Mohammad Sazzadul Hoque, Md. Abdul Mukit, Md. Abu Naser Bikas.
\newblock An Implementation of Intrusion Detection System Using Genetic Algorithm.
\newblock \emph{CoRR}, abs/1204.1336, 2012.
	\end{small}

>>>>

\begin{frame}
  \frametitle{The Big Picture}
  \begin{itemize}
  	\item Computer lab gets large numbers of login attempts that are attempts at intrusion.
	\item An attack could be trying to gain root access to the system.
	\item With an intrusion detection system (IDS), it would be possible to classify those attempts into legitimate and illegitimate attempts to login.
	\item Intrusion detection systems provide one way of detecting attacks by monitoring network activities for malicious or abnormal behaviors and then producing reports, alerts, and actions.
	\item A way of training an IDS about possible threats is by using a fuzzy genetic algorithm.
  \end{itemize}
\end{frame}

@@

\subsection{Detection Methodologies and Rules}
\begin{frame}
  \frametitle{Detection Methodologies}
  \begin{itemize}
  	\item Signature-based detection: compares well-known patterns of attacks
that are already known to the IDS against captured events in order to identify a possible attack.
	\begin{itemize}
		\item Simple and effective way to detect known attacks, but is ineffective against new kinds of unknown attacks.
	\end{itemize}

  	\item Anomaly-based detection: looks for patterns of activity that are rare and uncommon.
	\begin{itemize}
		\item Harder to do than signature-based detection, but it can be an effective way to detect new, unknown attacks.
	\end{itemize}
  \end{itemize}
\end{frame}

@@

\subsection{Data Sets - KDD99 and RLD09}
\begin{frame}
  \frametitle{KDD99}
	\begin{itemize}
		\item Generated by simulating a military network environment in 1999.
		\item Has long been a standard data set for intrusion detection.

		\item Data was processed into 5 million \emph{records}.
			\begin{itemize}
				\item A record is a sequence of TCP packets, between which data flows to and from a source IP address to a target IP address.
			\end{itemize}
		\item Data in the set is classified as normal or attack activity.

		\item Uses 41 \emph{features}, which are properties of a record that are used to describe the activity and help to distinguish normal connections from attacks.

@@

\begin{frame}
  \frametitle{RLD09}
	\begin{itemize}
		\item RLD09 was created because KDD99 is 14 years old, and newer attack types are not included in KDD99 because of its age.
		\item Data was captured from a university in Bangkok, Thailand.
		\item 17 different types of attacks (divided into denial of service and probe attacks).
	\end{itemize}
\end{frame}

@@

\subsection{Determining the Accuracy of an Algorithm}
\begin{frame}
  \frametitle{Determining the Accuracy of an Algorithm}
\centering{Predicted}
\begin{table}
Actual
\begin{tabular}{l|ll}
 & Not Attack & Attack \\ \hline
Not Attack & True Negative (TN) & False Positive (FP) \\
Attack & False Negative (FN) & True Positive (FP)  \\
\end{tabular}
\end{table}
\linespace
	\begin{itemize}
		\item Detection rate (DR): the percentage of normal and attack activity correctly classified from the total number of data records.
	\end{itemize}
\end{frame}

\subsection{Genetic Algorithms}
\begin{frame}
  \frametitle{Genetic Algorithms}
	\begin{itemize}
		\item GAs are a search technique used to find solutions to problems.
        \item Possible solutions to problems can be represented in a variety of problem dependent ways, such as bit strings.
        \item First, a randomly generated population of potential solutions is created. Then mutation, crossover, and selection are applied to each generation until an acceptable solution is found or some time limit is exceeded.
	\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Genetic Algorithms}
	\begin{itemize}
        \item Mutation: random bits in an individual, or possible solution, are randomly changed.
        \item Crossover: two individuals swap sequences of bits to form two new individuals.
        \item Selection: individuals that have better fitness are chosen to be parents.
        \item The fitness of an individual is specified by the fitness function, which determines the quality of a particular individual.
	\end{itemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Using Fuzzy Genetic Algorithms]{Using Fuzzy Genetic Algorithms}

@@

\begin{frame}
	\frametitle{Algorithm Overview}
	\begin{itemize}
		\item One record is passed into a rule.
		\item Each feature in a record is matched to one block of the rule.
		\item The parameters of each block measure the degree of certainty of an attack using the trapezoidal fuzzy rule shape.
		\item The sum of the degrees of certainty from each block are then compared with a threshold to determine if the record represents an attack or normal behavior.
	\end{itemize}
\end{frame}

@@

\begin{frame}
	\frametitle{Experiments}
	\begin{itemize}
		\item A variety of experiments were run. Two experiments used just RLD09. Three experiments used both RLD09 and KDD99 in order to compare how the fuzzy GA would perform on both.
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Experiments Using Only RLD09}
	Experiment 1
	\begin{itemize}
		\item Fuzzy GA was used to create DoS and probe detection rules. Both the DoS and probe rules were then used together in the testing process to identify attacks from the testing data set.
		\item 10,000 records in the training set, 26,500 records in the test set.
	\end{itemize}

@@

%	\bibitem{DBLP:journals/corr/abs-1204-1336}
%Mohammad Sazzadul Hoque, Md. Abdul Mukit, Md. Abu Naser Bikas.
%\newblock An Implementation of Intrusion Detection System Using Genetic Algorithm.
%\newblock \emph{CoRR}, abs/1204.1336, 2012.
	\end{small}
